### 2018-12-21 submission/manuscript.tex
Fix references

\setlength{\droptitle}{-2em}
  \title{}
  \pretitle{\vspace{\droptitle}}
  \posttitle{}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

@@

@@

\textbf{Background.} A typical 16S rRNA gene sequencing workflow can be
divided into preservation, extraction, amplification, and sequencing
steps. At each of these stages error can be introduced that will change
the underlying bacterial community composition results. In this study,
we focus on the amplification step's contribution to this overall error.
To accomplish this we assessed 16S rRNA gene sequencing results in mock
community and human fecal samples after using different amplification
cycles and high fidelity DNA polymerases.

\textbf{Methods.} Amplification of mock communities (technical
replicates n=4) consisting of previously isolated whole genomes of 8
different bacteria were amplifeid at 15, 20, 25, 30, and 35 cycles using
Accuprime, Kappa, Phusion, Platinum, or Q5 high fidelity DNA polymerase.
For fecal samples (n=4) we first extracted DNA using a PowerMag DNA
extraction kit with a 10 minute bead beating step and then used the same
approach as the mock community for the PCR amplification step. We first
assessed GC dependent differences, error rate, sequence error
prevalence, chimera prevalence, and correlation between chimera
prevalence and number of Operational Taxonomic Units (OTUs) by
polymerase and number of cycles. Next, differences in the number of OTUs
was examined based on the polymerase and number of cycles used.
Additionally, differences in the bacterial community composition was
assessed based on polymerase and number of cycles using the Bray-Curtis
index. Based on these findings individual taxa differences based on
polymerase and number of cycles was also investigated. Finally, Random
Forest models were created to test whether the bacterial community was
better at classifying polymerases, number of cycles, or individual
donor. We also assessed whether the most important taxa in the
polymerase and number of cycle Random Forest models were also important
in the model for individual donors.

\textbf{Results.} Predictably, we found noticeable differences in
relative abundance based on high and low GC content (P-value
\(\leqslant\) 0.04). Chimera prevalence in mock communities varied by
polymerase with differences being most notable at 35 cycles (Kappa =
5.71\% (median) versus Platinum = 26.62\%) and this variation persisted
after chimera removal using VSEARCH. We also observed positive
correlations between chimera prevalence and the number of OTUs with
Platinum having the highest (R\textsuperscript{2} = 0.974) and Kappa
having the worst (R\textsuperscript{2} = 0.478). When analyzing mock
community samples the variation in the number of OTUs detected by the
polymerases was observable as early as 20 cycles (P-value = 0.002). When
analyzing fecal samples we observed smaller differences in the number of
OTUs between polymerases. The median number of OTUs only varied by high
fidelity DNA polymerase used at 35 cycles and not at 20 cycles like the
mock community (P-value \textless{} 0.0001). Random Forest models were
most successful at classifying individual donor samples rather than
polymerase or number of cycles used (P-value \(\leqslant\) 5.49e-07).
Additionally, the most important OTUs in the polymerase and number of
cycle models were not the most important in the individual donor model.

\textbf{Conclusions.} Although there were 16S rRNA gene sequencing
differences based on polymerase and number of cycles used, bacterial
community composition differences were mostly only detectable in mock
communities. Collectively, these results provide evidence that smaller
biological differences between groups, based on 16S rRNA gene sequencing
of human fecal samples, can be consistently detected regardless of
polymerase and number of cycles used.

@@

The bacterial community is reported to vary between case and control for
a number of diseases (Turnbaugh et al., 2008; Sze et al., 2015; Baxter
et al., 2016; Bonfili et al., 2017). However, for diseases like obesity,
the taxa identified have varied widely depending on the study (Turnbaugh
et al., 2008; Zupancic et al., 2012). Although some of the differences
could be attributed to study design, error introduced during the 16S
rRNA gene sequencing workflow could also be another source of variation.
A typical 16S rRNA gene sequencing workflow can be divided into
preservation, extraction, PCR, and sequencing steps. The preservation
and extraction stages of the 16S rRNA gene sequencing workflow have been
the most extensively studied (Salter et al., 2014; Song et al., 2016;
Bassis et al., 2017; Kim et al., 2017). For preservation and extraction
stages of the workflow, it has been consistently found that there are
errors in the observed bacterial community composition based on the kits
used, but that these differences are smaller than the overall biological
difference measured between samples with different kits (Song et al.,
2016; Bassis et al., 2017). Since these studies use the same PCR
approach while varying preservation or extraction method, the
contribution of PCR error to this overall workflow is not as well
characterized.

There is a large body of literature that shows there are errors due to
primer and number of cycles chosen for the PCR stage of 16S rRNA gene
sequencing (Eckert \& Kunkel, 1991; Burkardt, 2000). Primers have
variable region dependent binding affinities which causes an inability
to detect specific bacteria (e.g.~V1-V3 does not detect
\emph{Haemophilus influenzae} and V3-V5 does not detect
\emph{Propionibacterium acnes}) (Sze et al., 2015 (Table S4); Meisel et
al., 2016). Another source of error is the selective amplification of
AT-rich over GC-rich sequences which exaggerate the difference between
16S rRNA genes higher in AT versus those higher in GC content (Polz \&
Cavanaugh, 1998). Many of these sources of error are made worse as the
number of cycles increases (Wang \& Wang, 1996; Haas et al., 2011;
Kebschull \& Zador, 2015). For example, both amplification error and
non-specific amplification (e.g.~incorrect amplicon size products) also
can increase as the number of cycles increases. This will increase the
number of Operational Taxonomic Units (OTUs) observed and drastically
change the values obtained from commonly used diversity measures (Acinas
et al., 2005; Santos et al., 2016). Additionally, as the number of
cycles increases more chimeras can form from an aborted extension step
that causes a priming error and subsequent secondary extension (Haas et
al., 2011). These chimeras will artificially increase community
diversity by increasing the number of OTUs that are observed (Haas et
al., 2011). In addition to these sources of errors, there also are
multiple families of DNA polymerases that have their own error rate and
proof reading capacity (Ishino \& Ishino, 2014). For simplicity, the use
of the term `error' within the context of this study will be used as a
broad term for changes in the downstream sequencing results due to the
amplification step (e.g.~higher GC content genomes, base substitutions,
chimeras, changes from the expected relative abundance, etc.).

Interestingly, the influence different DNA polymerases can have on the
observed 16S rRNA gene sequencing results have not been well studied
like some of the other sources of PCR-based error. A recent study found
differences in the number of OTUS and chimeras between normal and high
fidelity DNA polymerases (Gohl et al., 2016). The authors of this study
could reduce the difference between two polymerases by optimizing the
annealing and extension steps within the PCR protocol (Gohl et al.,
2016). Yet, within this study there was no comparison made between
different high fidelity DNA polymerases on the same samples. The authors
instead focused their efforts on testing specific types of amplification
methods. Additionally, the fecal samples that were analyzed, as a model
of a more complex community, may overestimate the biological variation
in human samples due to the comparison of captive versus semi-captive
red-shanked doucs. Due to these gaps, it makes sense to extend the
previous line of inquiry to include a more detailed examination of two
specific areas. First, whether differences in PCR error are detected
based on number of cycles used and high fidelity polymerases. Second,
whether these differences in error are enough to obscure smaller
biological signals than what has been previously reported (Gohl et al.,
2016).

This study will investigate whether there are error differences that are
dependent on high fidelity DNA polymerases and whether these differences
obscure the biological variation in the fecal bacterial communities
between individuals. We accomplish this by first examining GC-based
amplification error, sequence error (e.g.~substitutions), chimera
prevalence, the number of OTUs, Bray-Curtis index, and OTU-based
differences at varying number of cycles in five different high fidelity
DNA polymerases in mock communities. In addition to these metrics we
also created Random Forest models of the mock community to assess how
successful classification based on 16S rRNA gene sequencing of number of
cycles and type of polymerase could be. We then assessed if similar
differences in the number of OTUs, Bray-Curtis index, and OTUs could be
detected in human fecal samples. Additionally, we also created Random
Forest models to assess how successful they could be at classifying
fecal samples by number of cycles, polymerase used, or individual.
Collectively, our observations suggest that although detectable
differences occur based on number of cycles or high fidelity DNA
polymerase used, they are smaller than the biological differences
between individuals.

@@

\textbf{\emph{Error differences due to number of cycles or polymerase
used are detectable in mock communities.}} There was a significant
difference in relative abundance between high/low GC content based on
either the V4 16S rRNA gene region or the whole genome {[}Figure 1 \&
Table S1{]}. However, using only the GC content of the V4 region
resulted in less differences than when using the whole genome {[}Table
S1{]}. The differences between high and low GC content groups were
significant as early as 20 cycles in all polymerases except Q5 {[}Table
S1{]}. The highest (\emph{Staphylococcus}) and lowest
(\emph{Pseudomonas}) GC content bacteria were the most divergent from
the expected relative abundance of 0.12 {[}Figure 1{]}. These
observations suggest that the high fidelity DNA polymerases tested do
not prevent GC-based amplification error.

The median error rate varied by polymerase, with Kappa having the
highest error rate of all the polymerases across the number of cycles
{[}Figure 2A \& Table S2{]}. Thus, the majority of the differences
observed across the number of cycles was between Kappa and other
polymerases {[}Figure 2A \& Table S3{]}. The total sequences that
contained at least one error was also polymerase dependent, with the
majority of differences being between Kappa or Accuprime and the other
polymerases {[}Table S2 \& S3{]}. These differences in error rates were
not due to polymerase dependent differences in base substitution rate
{[}Figure S1{]}. Collectively, the results suggest that sequence error
is dependent on polymerase, persists across the number of cycles used,
and are not due to any bias towards a specific type of base
substitution.

We next examined whether chimeras also were dependent on polymerase and
whether this could affect the total number of OTUs. We observed
significant differences in the chimera prevalence based on polymerase
across the number of cycles used (P-value \textless{} 0.05) {[}Table
S2{]}. Differences in chimera prevalence between Platinum and all other
polymerases accounted for the majority of these differences {[}Table
S3{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
of all polymerases regardless of whether chimera removal with VSEARCH
was used {[}Figure 2B \& 2C{]}. The number of cycles used clearly
increased chimera prevalence for all polymerases but the rate of
increase differed {[}Figure 2B \& 2C{]}. There was also a plateau in the
total percent of chimeras that were removed that was similar for all
polymerases {[}Figure 2D{]}. Additionally, a positive correlation was
observed between chimeric sequences and the number of OTUs for all
polymerases {[}Figure 2E{]}. This positive correlation was strongest for
Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 2E{]}.
This data suggests that chimera prevalence depends on polymerase, is
made worse by increasing the number of cycles, and confirms previous
reports that the number of OTUs is dependent on the prevalence of these
chimeric sequences.

\textbf{\emph{Bacterial community composition differences based on
number of cycles or polymerase used are less obvious in mock community
samples.}} Due to the close relationship between chimera prevalence and
the number of OTUs, we also observed a polymerase dependent difference
in the range of the number of OTUs in the mock community samples based
on number of cycles and polymerase {[}Figure 3A{]}. The closest any of
the polymerases came to a total of 8 OTUs, created by running the mock
reference 16S sequences through mothur processing, was at 25 and 30
cycles {[}Figure 3A{]}. Differences between polymerases for the number
of OTUs created were observed as early as 20 cycles in the mock
community (F-stat = 15.82, P-value = 0.002) {[}Table S4{]}. Using a
Tukey post-hoc test, the majority of differences for the number of OTUs
detected in the mock community was largely due to Kappa and Platinum
versus the other polymerases across the number of cycles used {[}Table
S5{]}.

We next investigated the effect polymerases and number of cycles had on
the bacterial community composition in more detail by using the
Bray-Curtis index. We observed clear differences in the mock community
based on polymerase used (P-value = 1e-04) as well as differences based
on polymerase and number of cycles (P-value = 1e-04). However, when
investigating specific cycles within each polymerase we observed no
difference in the bacterial community composition (P-value
\textgreater{} 0.05) {[}Figure 3B{]}. Finding specific differences in
overall bacterial community composition based on number of cycles and
polymerase, we next examined if specific taxa vary depending on
polymerase and number of cycles. We found that OTUs that taxonomically
classified to \emph{Salmonella}, \emph{Escherichia},
\emph{Enterococcus}, and \emph{Staphylococcus} differed based on
polymerase used {[}Table S6 \& S7{]}. These specific OTUs were all
members of the mock community and the differences were most pronounced
at 25 and 30 cycles {[}Table S7{]}. The majority of the differences
between taxa were due to differences between a single polymerase (Kappa)
and the other polymerases tested {[}Table S7{]}.

Based on these observations, we created Random Forest models to classify
samples based on number of cycles or polymerase used. When assessing the
Random Forest models there was a difference between models built to
classify number of cycles versus those built to classify polymerase used
{[}P-value = 5.06e-07{]}. Although there was a difference, neither model
was overly successful at correctly classifying samples based on 16S rRNA
gene sequencing data (cycles model, probability of correct
classification = 0.42 (0.4 - 0.43)(min - max) versus polymerase model,
probability of correct classification = 0.39 (0.37 - 0.4)). Our
observations in mock communities suggest that although differences are
detected based on number of cycles or polymerase used, the overall
community-wide differences are not as large.

\textbf{\emph{Few differences based on number of cycles and polymerase
used were identified in fecal samples.}} Unlike mock community samples,
an overall difference in the number of OTUs was only detected between
polymerases at 35 cycles in fecal samples (F-stat \textgreater{} 16.35,
P-value = 9.7e-05) {[}Table S4{]}, and a Tukey post-hoc test failed to
identify any differences between specific polymerases used (P-value
\textgreater{} 0.05) {[}Table S5{]}. However, differences in the range
of the number of OTUs detected for fecal samples was dependent on the
polymerase used (e.g.~Accuprime at 35 cycles = 84 - 106 versus Phusion
at 35 cycles = 84 - 136) {[}Figure 4A{]}. There was also a trend for
lower number of cycles (15-20) to result in a reduced range in the
number of OTUs versus higher number of cycles (25, 30, and 35) for all
polymerases (e.g.~Phusion at 15 cycles = 10 - 19 versus Phusion at 35
cycles = 84 - 136) {[}Figure 4A{]}. Based on these observations, there
are small differences in the total number of OTUs detected in fecal
samples that are based on number of cycles or polymerases used.

When using the Bray-Curtis index, we observed that polymerases and
number of cycles had no affect on the bacterial community composition in
fecal samples (P-value = 1). There was also no interaction between
individuals, number of cycles, and polymerase used (P-value = 1). When
using PERMANOVA to test for community differences based on any of the
number of cycles within polymerases, only Phusion had cycle dependent
differences (P-value = 0.03). Additionally, Phusion was one of two
polymerases that had enough sequences to be rarefied to 1000 sequences
at 15 cycles. Within each respective 5-cycle increment comparison there
was no difference in Bray-Curtis index between the polymerases (P-value
\textgreater{} 0.05) {[}Figure 4B{]}. However, there was an overall
decrease in Bray-Curtis index by cycle comparison group (e.g.~the 15
cycle versus 20 cycle group compared to the 30 cycle versus 35 cycle
group) (P-value \textless{} 0.01) {[}Figure 4B{]}. Using a Dunn's
post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle
comparison groups had a higher Bray-Curtis index than the 30 cycle
versus 35 cycle comparison group (P-value \textless{} 0.05). In addition
to assessing community composition by Bray-Curtis index, we also
investigated differences in OTUs based on polymerase at each number of
cycle used and no significant differences could be found (P-value
\(\geqslant\) 0.01, corrected P-value = 1). Overall, these results
suggest that in fecal samples the number of cycles and polymerase used
minimally change the bacterial community composition.

Finally, we built Random Forest models to classify samples for the
number of cycles, polymerase used, or individual it was obtained from.
Using 16S rRNA gene sequencing data there was a large disparity in the
model's ability to correctly classify samples based on polymerase used,
number of cycles, and individual (P-value \(\leqslant\) 5.49e-07). There
was a clear difference in the success of the model used to classify
individuals (probability of correct classification = 0.87 (0.86 - 0.9))
versus the models created to classify number of cycles or polymerase
(cycles model, probability of correct classification = 0.26 (0.25 -
0.27)(min - max) and polymerase model, probability of correct
classification = 0.18 (0.17 - 0.2)). Additionally, the top 10 most
important OTUs to the polymerase and number of cycle models were not the
most important OTUs to the model that classified individuals {[}Figure
5{]}. This suggests that OTUs that change the most due to the number of
cycles and polymerase in fecal samples are not the same as OTUs that are
part of the biological variation between individuals {[}Figure 5{]}.
Collectively, our observations show that despite a variety of errors
associated with the amplification process, it is still possible to
detect the bacterial community variation between individuals.

@@

Previous studies have reported GC dependent amplification differences
(Polz \& Cavanaugh, 1998) and we confirm that these differences still
persist in high fidelity DNA polymerase when analyzing mock community
samples {[}Figure 1{]}. Additionally, these specific polymerases have
different sequence error rates and chimera prevalence {[}Figure 2{]}.
Chimera prevalence in particular was strongly associated with the total
number of OTUs detected in our mock community and vaired by polymerase
{[}Figure 2E \& 3A{]}. Additionally, the bacterial community composition
varied by number of cycles and polymerase in these mock community
samples. Clearly, it is possible to identify a large number of
differences based on number of cycles and polymerase used when analyzing
mock community samples. However, these differences may not be a major
concern when analyzing more complex community samples.

Fecal samples showed a similar increase in median and range for the
number of OTUs, based on number of cycles and polymerase used, as that
found when analyzing mock community samples {[}Figure 4A{]}. However,
there were few differences between the polymerases within each of the
amplification cycles {[}Figure 4A{]}. Additionally, these differences
did not translate to large bacterial community composition changes based
on number of cycles or polymerase. The differences that were identifed
were small and occured based on the number of cycles, independent of the
polymerase used {[}Figure 4B{]}. There was also no OTU-based differences
observed within the number of cycles between any of the polymerases for
fecal samples. Interestingly, Random Forest models built on this OTU
data had the highest classification success for individuals versus
number of cycles or polymerase. In addition to the better model
performance of Random Forest for individuals, the most important OTUs to
this model were not the most important OTUs for the other models tested
{[}Figure 5{]}. Collectively, these observations show that differences
based on the number of cycles and between high fidelity DNA polymerases
can be found. However, when using 16S rRNA gene sequencing these
technical differences can be smaller than the biological differences
between fecal bacterial communities within individuals.

>>>>

\setlength{\droptitle}{-2em}

  \title{}
    \pretitle{\vspace{\droptitle}}
  \posttitle{}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

@@

\section{Influence of DNA Polymerase on PCR bias, chimera formation, and
errors when sequencing 16S rRNA genes from complex
communities}\label{influence-of-dna-polymerase-on-pcr-bias-chimera-formation-and-errors-when-sequencing-16s-rrna-genes-from-complex-communities}

@@

@@

16S rRNA gene sequencing is a powerful and widely used tool for
surveying the structure of microbial communities (1--3). This approach
has exploded in popularity with the advent of high throughput sequencing
where it is possible to characterize numerous samples with thousands of
sequences per sample. Numerous factors can impact how a natural
community is represented by the sequencing data including the method of
acquiring samples (4--8), storage conditions (4--6, 9--12), extraction
methods (13), amplification conditions (8, 14, 15), sequencing method
(15--17), and analytical pipeline (15, 18--20). The increased sampling
depth that is now available relative to previous Sanger-based methods is
expected to compound the impacts of an investigator's choices and the
interpretation of their results.

One step in the generation of 16S rRNA gene sequence data that has been
long known to have a significant impact on the description of microbial
communities is the choice of conditions for PCR amplification (8, 14,
15). Factors such as the choice of primers have an obvious impact on
which populations will be amplified (18, 21). However, a variety of PCR
artifacts can also impact the perception of a community including the
formation of chimeras (14, 22--24), misincorporation of nucleotides (23,
25, 26), preferential amplification of populations leading to bias (24,
27--33), and accumulation of random amplification events leading to PCR
drift (24, 27, 32, 34). Many bioinformatic tools have been developed to
identify chimeras, there are significant sensitivity and specificity
tradeoffs (14, 35). Others have attempted to account for PCR bias using
modeling approaches (29, 36); however, these have been developed for
idealized situations. Laboratory-based solutions to minimize chimera
formation have also been proposed such as minimizing the amount of
template DNA in the PCR, minimizing the number of rounds of PCR,
minimizing the amount of shearing in the template DNA, and using DNA
polymerases that have a proof-reading ability (14, 23). To minimize PCR
drift, some investigators will pool technical replicate PCRs hoping to
average out the drift (34). Other factors that have been shown to impact
the formation of PCR artifacts are outside the control of an
investigator including the fraction of DNA bases that are guanines or
cytosines, the variation in the length of the targeted region across the
community, the sequence of the DNA that flanks the template, and the
genetic diversity of the community (28, 30--33). Early investigations of
the factors that lead to the formation of PCR artifacts focused on
analyzing binary mixtures of genomic DNA and 16S rRNA gene fragments to
explore PCR biases and chimera formation. Although these studies were
instrumental in forcing researchers to be cautious about the
interpretation of their results, we have a poor understanding of how
these factors affect the formation of PCR artifacts in more complex
communities.

The influence that the choice of DNA polymerase has on the formation of
PCR artifacts has not been well studied. There has been recent interest
in how the choice of the hypervariable region and data analysis
pipelines impact the sequencing error rate (15, 18--20); however, these
studies use the same DNA polymerase in the PCR step and implicitly
assume that the rate of nucleotide misincorporation from PCR are
significantly smaller than those from the sequencing phase. There has
been more limited interest in the impact that DNA polymerase choice has
on the formation of chimeras (23, 37). A recent study found differences
in the number of OTUs and chimeras between normal and high fidelity DNA
polymerases (37). The authors of this study could reduce the difference
between two polymerases by optimizing the annealing and extension steps
within the PCR protocol (37). Yet this optimization was specific for the
community they were analyzing (i.e.~captive and semi-captive red-shanked
doucs) and assumed that if the two polymerases generate the same
community structure that the community structure was correct. In fact,
the community structure generated by both methods was not free of
artifacts, but has the same artifacts. A challenge in these types of
experiments is having a priori knowledge of the true community
representation. A mock community with known composition allows
researchers to quantify the sequencing error rate, fraction of chimeras,
and bias (19); however, mock communities have a limited phylogenetic
diversity relative to natural communities. Natural communities, in
contrast, have an unknown community composition making absolute
measurements impossible. They can be used to validate results from mock
communities and to understand the relative impacts of artifacts on the
ability to differentiate biological and methodological sources of
variation. Given the large number of DNA polymerases available to
researchers it is unlikely that a specific recommendation is possible.
Rather, the development of general best practices and understanding the
impact of PCR artifacts on an analysis are desired.

This study investigated the impact of choice of high fidelity DNA
polymerase and the number of rounds of amplification on the formation of
PCR artifacts using a mock community and human stool samples. These PCR
artifacts included (i) the effect of the polymerase on the error rate of
the bases represented in the final sequences, (ii) the fraction of
sequences that appeared to be chimeras and the ability to detect those
chimeras, (iii) the bias of preferentially amplifying one fragment over
another in a mixed pool of templates, and (iv) inter-sample variation in
community structure of samples amplified with the same polymerase across
the amplification process. To characterize these factors we sequenced a
mock community of 8 organisms with known sequences and community
structure and human fecal samples with unknown sequences and community
structures. We sequenced the V4 region of the 16S rRNA genes from a mock
community by generating paired 250 nt reads on the Illumina MiSeq
platform. This region and sequencing approach was used because it has
been shown to result in a relatively low sequencing error rate and is a
widely used protocol. To better understand the impact of DNA polymerase
choice on PCR artifacts, we selected five high fidelity DNA polymerases
and amplified the communities using 15, 20, 25, 30, and 35 rounds of
amplification (Table 1). Collectively, our results suggest that the
number of rounds and to a lesser degree the choice of DNA polymerase
used in PCR impact the sequence data, the effects are consistent and are
smaller than the biological differences between individuals.

@@

\textbf{\emph{Sequencing errors vary by the number of cycles and the DNA
polymerase used in PCR.}} The presence of sequence errors can confound
the ability to accurately classify 16S rRNA gene sequences and group
sequences into operational taxonomic units (OTUs). More importantly,
sequencing errors themselves can alter the representation of the
community. Therefore, it is important to minimize the number of
sequencing errors. Using a widely-used approach that generates the
lowest reported error rate, we quantified the error rate by sequencing
the V4 region of the 16S rRNA genes from an 8 member mock community. We
also removed any contigs that were at least three bases more similar to
a chimera of two references than to a single reference sequence
{[}ref{]}. Regardless of the polymerase, the error rate increased with
the number of rounds of amplification. Using 30 rounds of PCR is a
common approach across diverse types of samples. Among the data
generated using 30 rounds of PCR the Accuprime polymerase had the
highest error rate (i.e.~0.XXX\%) followed by the Platinum
(i.e.~0.XXX\%), Phusion (i.e.~0.XXX\%), Kappa (i.e.~0.XXX\%), and Q5
(i.e.~0.XXX\%) polymerases (Figure 1). When we applied the
pre.clustering denoising step, which merges the counts of reads within 2
nt of a more abundant sequence, the error rates dropped considerably
such that the Platinum polymerase had the highest error rate
(i.e.~0.XXX\%) followed by the Accuprime (i.e.~0.XXX\%), Phusion
(i.e.~0.XXX\%), Kappa (i.e.~0.XXX\%), and Q5 (i.e.~0.XXX\%) polymerases
(Figure 1). Although specific recommendations are difficult to make
because the impact of the actual community structure and concentration
of the initial DNA template are likely to have an impact on the results,
it is clear that using as few PCR cycles as necessary and a polymerase
with the lowest possible error rate is a good guide to minimizing the
impact of polymerase on the error rate.

\textbf{\emph{The fraction of sequences identified as being chimeric
varies by the number of cycles and the DNA polymerase used in PCR.}}
Chimeric PCR products can significantly confound downstream analyses.
Although numerous bioinformatic tools exist to identify and remove
chimeric sequences with high specificity, their sensitivity is
relatively low and can be reduced by the presence of sequencing errors.
We identified those contigs that were at least three bases more similar
to a chimera of two references than to a single reference sequence. As
expected from previous studies, the number of chimeras increased with
rounds of amplification. Interestingly, the fraction of chimeric
sequences from the mock community varied by the type of polymerase used.
After 30 rounds of PCR, the Platinum polymerase had the highest chimera
rate (i.e.~XX\%) followed by the Q5 (i.e.~XX\%), Phusion (i.e.~XX\%),
Kappa (i.e.~XX\%), and Accuprime (i.e.~XX\%) polymerases. We used the
UChime algorithm to detect chimeras in our pre-clustered mock community
data and calculate the algorithm's sensitivity and specificity. For all
polymerases, the specificity was above XX\% and showed a weak
association with the number of cycles used. There was considerable
inter-polymerase and inter-round of amplification variation in the
sensitivity of UChime to detect the chimeras from the mock community.
This suggested that the residual error rate after pre-clustering the
sequence data did not compromise the sensitivity. In general, the
sensitivity of UChime varied between XX and XX\%. The generalizability
of these results is limited because we used a single mock community with
limited genetic diversity. Although we did not know the true chimera
rate for our four human stool samples, we were able to calculate the
fraction of sequences that UChime identified as being chimeric. These
results followed those from the mock communities; additional rounds of
amplification significantly increased the rate of chimeras and there was
variation between the polymerases that we used. Although it was not
possible to identify the features of a polymerase that resulted in
higher rates of chimeras, it is clear that using the smallest number of
PCR cycles possible will minimize the impacts of chimeras.

\textbf{\emph{At the sequence level, PCR amplification bias is subtle.}}
Since researchers began using PCR to sequence 16S rRNA genes there has
been concern that amplifying fragments from a mixed template pool could
lead to a biased representation in the pool of products and would
confound downstream analyses. The mock community was generated by mixing
equal amounts of genomic DNA from each of the 8 bacteria resulting in
uneven representation of the \emph{rrn} operons across the bacteria. The
vendor of the mock community subjects each lot of genomic DNA to shotgun
sequencing to more accurately quantify the actual abundance of each
organism in the community. We compared the expected relative abundance
of the 16S rRNA genes from each bacterium in the mock community to the
data we generated across rounds of amplification and polymerase.
Interestingly, for some bacteria, their representation became less
biased with additional rounds of PCR (e.g. \emph{L. fermentum}), while
others became more biased (e.g. \emph{E. faecalis}), and others had
little change (e.g. \emph{B. subtilis}). The percentage of bases in the
V4 region that were guanines or cytosines was not predictive of the
amount of bias. Across the strains there was no variation in the length
of their V4 regions and they each had the same sequence in the region
that the primers annealed. One of the bacteria represented in the mock
community, \emph{S. enterica}, had 6 identical copies of the V4 region
and 1 copy that differed from those by one nucleotide. The dominant copy
had a thymidine and the rare copy had a guanine. We used the sequence
data to calculate the ratio of the dominant to rare variants from
\emph{S. enterica} (Figure. XXX). The Accuprime, Phusion, Platinum, and
Q5 polymerases converged to a ratio of 5.XXX; however, the ratio for the
Kappa polymerase varied between 6.XXX and 7.XXX for 25 to 35 rounds of
PCR. Given the subtle nature of the variation in the relative abundances
of each 16S rRNA gene fragment, it was not possible to create
generalizable rules that would explain the bias and the previous
observation regarding the guanine and cytosine relative abundance was
not observed with our data.

\textbf{\emph{At the community level, the effects of PCR amplification
bias grow with additional rounds of PCR.}} Because the distance between
samples and across rounds of PCR could be artificially inflated due to
sequencing errors and chimeras, we used analyzed the alpha and beta
diversity of the mock community data at different phases of the sequence
curation pipeline. First, we removed the chimeras from the mock
community data as described above and mapped the individual reads to the
OTUs that the 16S rRNA gene fragments would cluster into if there were
no sequencing errors. This gave us a community distribution that
reflected the distribution following PCR without any artifacts. Alpha
diversity metrics such as richness and Shannon and Inverse Simpson
diversity metrics increased with the number of rounds of PCR for all
polymerases except for the Kappa polymerase, where the richness remained
constant but the diversity decreased. These data suggest that PCR had
the effect of making the community distribution more even than it was
originally. Next, we used the true sequence diversity, but removed
chimeras based on mapping reads to all possible chimeras between
reference sequences, and clustered the reads to OTUs. The same trends
were observed as with perfect sequence data except that the richness and
diversity metrics trended higher. Finally, we used the observed sequence
data and used the UCHIME algorithm to identify chimeras. Again, we
observed the same trends as with the other analyses except the richness
and diversity metrics trended higher. These comparisons demonstrated
that although the bias at the sequence level was subtle, PCR does have a
bias that is exacerbated by errors and chimeras. When we measured the
Bray-Curtis distance between the communities observed after 25 rounds of
amplification and those at 30 and 35 rounds of amplification the
distances between 25 and 35 rounds were higher than between 25 and 30
rounds for each of the polymerases by an average of 0.XX units. The
Platinum polymerase varied the most across rounds of amplification (25
vs 30 rounds: 0.XX; 25 vs 35 rounds: 0.XX). Although the distances
between samples were small, the ordination of these distances showed a
clear change in community structure with increasing rounds of PCR. This
observation was supported by our statistical analysis, which revealed
that the number of rounds of PCR (R\textsuperscript{2}=0.XX, P=0.XXXX)
was a larger factor than the choice of polymerase
(R\textsuperscript{2}=0.XX, P=0.XXXX). These results demonstrate that
subtle differences in relative abundances can have an impact on overall
community structure. This variation underscores the importance of only
comparing sequence data that have been generated using the same PCR
conditions.

\textbf{\emph{The choice of polymerase or the number of rounds of
amplification have little impact on the relative interpretation of
community-wide metrics of diversity.}} The biases that we observed at
the population and community levels using mock community data appeared
to be small relative to the expected differences between biological
samples. To study this further, we calculated alpha and beta-diversity
metrics using the human stool samples for each of the polymerases and
rounds of amplification. We calculated the number of observed OTUs,
Shannon diversity, inverse Simpson diversity index for each condition
and donor. Although there were clear differences between conditions, the
relative ordering of the stool samples did not meaningfully vary across
conditions. When we characterized the variation between rounds of
amplification using human stool samples, the distance between the 25 and
30 rounds and 25 and 35 rounds varied considerably between samples and
polymerases. In general the inter-round variation was lowest for the
data generated using the Kappa and Accuprime polymerases. The data
generated using the Platinum polymerase was consistent across rounds,
but it was more biased than these polymerases. Considering the average
distance across the four samples varied between 0.XX and 0.XX,
regardless of the polymerases and number of rounds of amplification, any
bias due to amplification is unlikely to obscure community-wide
differences between samples. In support of this was our principle
coordinates analysis of the Bray-Curtis distances, which revealed
distinct clusters by donor. Within each cluster there were no obvious
patterns related to the polymerase or number of rounds of PCR. However,
our statistical analysis using adonis revealed statistically significant
differences in the community structures with the subject explaining the
most variation (R\textsuperscript{2}=0.XXX, P=0.XXXX), followed by the
number of rounds of PCR (R\textsuperscript{2}=0.XXX, P=0.XXXX) and the
choice of polymerase (R\textsuperscript{2}=0.XXX, P=0.XXXX). Together,
these results indicate that for a coarse analysis of communities, the
choice of number of rounds of amplification or polymerase are not
important, but that they must be consistent across samples. It is
difficult to develop a specific recommendation based on the level of
bias across rounds of PCR or polymerases; however, the general
suggestion is to use as few rounds of amplification as possible.

\textbf{\emph{There is little evidence of a relationship between
polymerase or number of rounds of amplification on PCR drift.}} There
have been concerns that the same template DNA subjected to the same PCR
conditions could result in different representations of communities
because of random drift. To test this, we determined the average
Bray-Curtis distance between replicate reactions using the same
polymerase and number of rounds of amplification. Using the mock
community data there were no obvious trends. The average Bray-Curtis
distance within a set of conditions varied by 0.XX to 0.XX units.
Although we did not obtain replicates of each of the stool samples, the
intra-sample variation for each set of conditions was consistent and
varied between 0.XX and 0.XX units. These data suggest that amplicon
sequencing is robust to random variation in amplification and that any
differences are likely to be smaller than what is considered
biologically relevant.

@@

Our results suggest that the number of rounds of PCR and to a lesser
degree the choice of DNA polymerase impact the analysis of 16S rRNA gene
sequence data from bacterial communities. Although it was not possible
to make direct connections between PCR conditions and specific sources
of bias, we were able to identify general recommendations that reduce
the amount of error, chimera formation, and bias. Researchers should
strive to minimize the number of rounds of PCR and should use a high
fidelity polymerase. Although specific PCR conditions impact the precise
interpretation of the data, the effects were consistent and were smaller
than the biological differences between individuals. Based on these
observations, within a study, amplicons must be generated by consistent
protocols to yield meaningful comparisons. When comparing across
studies, values like richness, diversity, and relative abundances must
be made in relative and not absolute terms.

The observed sequencing error rates and alpha diversity metrics followed
the manufacturers' measurements of their polymerases' fidelity.
Accuprime and Platinum have fidelity that are approximately 10-times
higher than that of Taq whereas the fidelity of Phusion, Q5, and Kappa
are more than 100 times higher. Among these polymerases, the Kappa
polymerase resulted in the the lowest error rate, lowest chimera rate,
and least bias across rounds of PCR. Considering polymerase development
is an active area of commercial development and it likely that new
polymerases will come to market, it is important for researchers to
understand how changing the polymerase impacts downstream analyses for
their type of samples.
